Quantifying Limits to Detection of Early Warning for Critical Transitions

### 2012-01-19 boettiger.tex
writing

\textbf{arbitrary windows}
While superficially simple to use, the calculation of statistics is subject to 
several somewhat arbitrary decisions and additional assumptions,
such as the choice of window size over which the statistic is computed,
the choice of detrending,
whether the consecutive averaging windows should overlap.  
\citet{Lenton2012} compare the performance of different approaches to these issues,
showing that choice of window and method used for detrending indeed influence the results,
and each approach considered has pros and cons.  
Without a consistent choice or a clear mechanism to determine one, % Should we say this part?
it is difficult to evaluate and compare between such methods.  
A model-based approach side-steps these issue entirely.  
% discusses these challenges, as does most of the others to lesser extent 
% Should we mention the ergodic problem here - that averaging over windows isn't averaging across time series, 
% and that it assumes stationarity, which is precisely the thing we're looking to change?  
% Similarly the detrending approaches can remove the signal?
% Should we also mention that some popular approaches such as autocorrelation require evenly-spaced points, and interpolating is incorrect, introducing fictitious correlations?

2) The warning signal, ``an increase in statistic $x$,'' is not quantitative,
making it difficult to compare between signals or to
attribute a statistical significance to the detection.
While much of the literature employing these statistics simply avoids 
any attempt to quantify the significance of the detection, 
\citet{Dakos2008} relies on Kendall's correlation coefficient to quantify an increase,
and then goes through some pains to consider potential null models
that could be used to calculate statistical significance.  
Other measures of increase, such as Pearson's, have also been proposed.

3) Specifying an appropriate null model is difficult.  
% simulation of replicates under the original model
As the summary statistic approach 
the non-parametric significance test for Kendall's tau tests against the null model that the sequence is independent,
something that is clearly violated when the statistic is computed in a sliding window.
Not surprisingly, \citet{Dakos2008} find this method overestimates the number of significant events relative to other nulls.  
Other non-parametric methods are tempting, 
such as shuffling the order of the data-points in the time-series,
but this breaks even the natural autocorrelation expected in a stable system that is not approaching any transition,
and thus should not be considered for a null.  
Relying on a more explicit but flexible model %or class of models
such as autogregressive processes is more consistent.  
% Does this go here?
We will see that such generic model-based approaches are becoming more common, 
and argue how they can be used to maximum advantage
% by simulation, likelihood fitting, and ROC curves.

@@

4) The underlying assumption that the system contains a saddle-node bifurcation
can be easilyoverlooked in summary-statistic based approach.  
For instance, variance may increase for a wide variety of reasons that 
do not signal an approaching transition~\citep{Schreiber2003, Schreiber2008} 
or may not increase as a bifurcation is approached~\citep{Livina2012, Dakos2011a}.

% Does this section go here?
A system may experience a critical transition due to a rare fluctuation,
or a perturbation to the system state, %(Does this need a figure)?
rather than a bifurcation~\citep{Scheffer2001, Scheffer2009}.  
When applying detection approaches to data that have already experienced a critical transition,
we must condition on the 
and attempting to ascertain the cause of the transition
These are alternative null hypotheses 
 %(Dakos 2008, Lenton,  see Lenton 2009 transition, see 
\citep{Ditlevsen2010} % telling these two cases apart
\citep{Livina2012} noise induced transitions cannot be forecast.

>>>>

\textbf{arbitrary windows}\\
While superficially simple to use, the calculation of statistics is subject to 
several somewhat arbitrary decisions and additional assumptions,
such as the choice of window size over which the statistic is computed,
the choice of detrending,
whether the consecutive averaging windows should overlap.  
\citet{Lenton2012} compare the performance of different approaches to these issues,
showing that choice of window and method used for detrending indeed influence the results,
and each approach considered has pros and cons.  
Without a consistent choice or a clear mechanism to determine one, % Should we say this part?
it is difficult to evaluate and compare between such methods.  
A model-based approach side-steps these issue entirely.  
% discusses these challenges, as does most of the others to lesser extent 
% Should we mention the ergodic problem here - that averaging over windows isn't averaging across time series, 
% and that it assumes stationarity, which is precisely the thing we're looking to change?  
% Similarly the detrending approaches can remove the signal?
% Should we also mention that some popular approaches such as autocorrelation require evenly-spaced points,
% and interpolating is incorrect, introducing fictitious correlations?

\textbf{no quantitative measures}\\
The warning signal, ``an increase in statistic $x$,'' is not quantitative,
making it difficult to compare between signals or to
attribute a statistical significance to the detection.
While much of the literature employing these statistics simply avoids 
any attempt to quantify the significance of the detection, 
\citet{Dakos2008} relies on Kendall's correlation coefficient to quantify an increase,
and then goes through some pains to consider potential null models
that could be used to calculate statistical significance.  
Other measures of increase, such as Pearson's, have also been proposed.

\textbf{problem of null models}\\
Specifying an appropriate null model is difficult.  
% simulation of replicates under the original model
As the summary statistic approach 
the non-parametric significance test for Kendall's tau tests against the null model that the sequence is independent,
something that is clearly violated when the statistic is computed in a sliding window.
Not surprisingly, \citet{Dakos2008} find this method overestimates the number of significant events relative to other nulls.  
Other non-parametric methods are tempting, 
such as shuffling the order of the data-points in the time-series,
but this breaks even the natural autocorrelation expected in a stable system that is not approaching any transition,
and thus should not be considered for a null.  
Relying on a more explicit but flexible model %or class of models
such as autogregressive processes is more consistent.  
% Does this go here?
We will see that such generic model-based approaches are becoming more common, 
and argue how they can be used to maximum advantage
% by simulation, likelihood fitting, and ROC curves.

@@

\textbf{hidden assumptions}\\
The underlying assumption that the system contains a saddle-node bifurcation
can be easily overlooked in summary-statistic based approach.  
For instance, variance may increase for reasons that 
do not signal an approaching transition~\citep{Schreiber2003, Schreiber2008}.
Alternatively, variance may not increase as a bifurcation is approached~\citep{Livina2012, Dakos2011a}.
Some classes of sudden transitions may exhibit no warning signals~\cite{Hastings2010}.

Even when the system is characterized by 
simple alternative stable states in the saddle-node model,
transitions between states may be due to other mechanisms than a bifurcation, such as a purely noise-induced transition,
or a

% Does this section go here?
A system may experience a critical transition due to a rare fluctuation,
or a perturbation to the system state, %(Does this need a figure)?
rather than a bifurcation~\citep{Scheffer2001, Scheffer2009}.

When applying detection approaches to data that have already experienced a critical transition,
we must condition on the

and attempting to ascertain the cause of the transition
These are alternative null hypotheses 
 %(Dakos 2008, Lenton,  see Lenton 2009 transition, see 
\citep{Ditlevsen2010} % telling these two cases apart
\citep{Livina2012} noise induced transitions cannot be forecast.
